The great codivergence bake off

### 2024-10-08 manuscript/ms.tex
Working on methods and results.

@@

>>>>

We simulated loci for 5 diploid individuals (i.e., 10 gene
copies) per population from 5 pairs of diverging populations.
Each pair shared a common ancestral population with a constant effective size
\epopsize[\rootpopindex]
that diverged at time
\comparisonetime{}
into two descendant populations with constant effective sizes
\epopsize[\descendantpopindex{1}]
and 
\epopsize[\descendantpopindex{2}].
Each locus we simulated comprised 500 linked characters with no intralocus
recombination.
We simulated datasets comprising 1, 2, 5, and 20 independent loci.
For each number of loci, we simulated datasets under three divergence
scenarios:
\begin{enumerate}
    \item All five pairs of populations diverged independently.
    \item All five pairs of populations diverged simultaneously.
    \item The number of divergence times (\nevents) and the assignment of the
        five pairs of populations to the divergene times (\divtimesets) vary
        according to a Dirichlet-process model with a concentration parameter
        of 2.22543.
\end{enumerate}
For each of the 12 simulation conditions (4 numbers of loci $\times$ 3
divergence scenarios),
we simulated 1000 datasets.
Each unique divergence time was drawn from an exponential distribution,
$\comparisonetime \sim \dexponential{0.01}$.
We constrained the mutation rate (\murate{}) to 1 across the five pairs of
populations, such that divergence times and effective population sizes were
scaled by the mutation rate.

\subsubsection{Full-likelihood analyses}

To simulate datasets and analyze them under a full-likelihood Bayesian model of
population divergences, we used the \simcoevolity and \ecoevolity tools of the
\ecoevolity software package
\citep[Version 0.3.2, Commit c128046;][]{Oaks2018ecoevolity,Oaks2018paic}.

The mutation-scaled effective sizes of the descendant populations
($\epopsize\murate$) were drawn from a gamma distribution with a shape of 10
and mean of 0.0005,
$\epopsize{}\murate{} \sim \dgamma{10}{0.0005}$.
The effective size of the ancestral population relative to the mean of
its descendants was drawn from a gamma distribution with a shape of 40 and mean of 1,
$\rootrelativepopsize \sim \dgamma{40}{1.0}$.

When analyzing each simulated \dataset with \ecoevolity, we ran four MCMC
chains for 30,000 generations sampling every 20 generations.
We used the last 1000 samples from each of the four chains to approximate the
posterior distribution of the model given each simulated dataset.
To assess convergence and mixing of the MCMC chains, we calculated the
potential scale reduction factor \citep[PSRF; the square root of Equation 1.1
in][]{Brooks1998} and effective sample size \citep[ESS;][]{Gong2014} for all
continuous parameters and the log likelihood using
\pycoevolity
\citep[Version 0.2.11 Commit 85ea44b;][]{Oaks2018ecoevolity,PycoevolityRepoOnline}.
% When plotting results, we highlight any simulation replicates that have a
% $\textrm{PSRF} > 1.2$.

\subsubsection{ABC analyses}

To simulate datasets and analyze them under a similar model estimated via
approximate-Bayesian computation (ABC),
we used \pymsbayes
\citep[Version 0.3.7, Commit 54f720df;][]{Oaks2014dpp},
which is a Python multiprocessing wrapper around dpp-msbayes
\citep{Oaks2014dpp}, which is a modification of
MTML-msBayes \citep{Huang2011}.
In this software, the distribution for effective population sizes is placed
on the parameter $4\epopsize{}\murate{}$, for which we used
a gamma distribution with a shape of 10 and a mean of 0.002.
Note, this is equivalent to the distribution we used for $\epopsize{}\murate{}$
in \ecoevolity.

@@

The potential scale reduction facter (PSRF) was less than X and the effective
sample size (ESS) was greater than Y for all parameters and the log likelihood
across all simulations
(Fig.\ S\labelcref{fig:psrfdivtime,fig:psrfrootpopsize,fig:psrflnl,fig:essdivtime,fig:essrootpopsize,fig:esslnl}).

### 2024-10-08 manuscript/si-figures.tex
Working on methods and results.

>>>>

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=0.9\textheight,keepaspectratio]{../images/from-project-repo/plots/tex-plot-grids/grid-psrf-div-time-histograms-cropped.pdf}
        \caption{
            The potential scale reduction factor \citep[PSRF; the square root
            of Equation 1.1 in][]{Brooks1998}
            for the divergence time parameter calculated from the last 1000
            samples of the four MCMC chains run for each simulation replicate.
            The mean PSRF across the 1000 simulation replicates is shown at the
            top right of each plot, followed by the range in parentheses.
        }
        \label{fig:psrfdivtime}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=0.9\textheight,keepaspectratio]{../images/from-project-repo/plots/tex-plot-grids/grid-psrf-root-pop-size-histograms-cropped.pdf}
        \caption{
            The potential scale reduction factor \citep[PSRF; the square root
            of Equation 1.1 in][]{Brooks1998}
            for the ancestral population size calculated from the last 1000
            samples of the four MCMC chains run for each simulation replicate.
            The mean PSRF across the 1000 simulation replicates is shown at the
            top right of each plot, followed by the range in parentheses.
        }
        \label{fig:psrfrootpopsize}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=0.9\textheight,keepaspectratio]{../images/from-project-repo/plots/tex-plot-grids/grid-psrf-lnl-histograms-cropped.pdf}
        \caption{
            The potential scale reduction factor \citep[PSRF; the square root
            of Equation 1.1 in][]{Brooks1998}
            for the log likelihood calculated from the last 1000
            samples of the four MCMC chains run for each simulation replicate.
            The mean PSRF across the 1000 simulation replicates is shown at the
            top right of each plot, followed by the range in parentheses.
        }
        \label{fig:psrflnl}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=0.9\textheight,keepaspectratio]{../images/from-project-repo/plots/tex-plot-grids/grid-ess-div-time-histograms-cropped.pdf}
        \caption{
            The effective sample size \citep[ESS;][]{Gong2014} for the
            divergence time parameter calculated from the last 1000 samples of
            the four MCMC chains run for each simulation replicate.
            The mean ESS across the 1000 simulation replicates is shown at the
            top right of each plot, followed by the range in parentheses.
        }
        \label{fig:essdivtime}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=0.9\textheight,keepaspectratio]{../images/from-project-repo/plots/tex-plot-grids/grid-ess-root-pop-size-histograms-cropped.pdf}
        \caption{
            The effective sample size \citep[ESS;][]{Gong2014} for the
            ancestral population size calculated from the last 1000 samples of
            the four MCMC chains run for each simulation replicate.
            The mean ESS across the 1000 simulation replicates is shown at the
            top right of each plot, followed by the range in parentheses.
        }
        \label{fig:essrootpopsize}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=0.9\textheight,keepaspectratio]{../images/from-project-repo/plots/tex-plot-grids/grid-ess-lnl-histograms-cropped.pdf}
        \caption{
            The effective sample size \citep[ESS;][]{Gong2014} for the
            log likelihood calculated from the last 1000 samples of
            the four MCMC chains run for each simulation replicate.
            The mean ESS across the 1000 simulation replicates is shown at the
            top right of each plot, followed by the range in parentheses.
        }
        \label{fig:esslnl}
    \end{center}
\end{figure}
